Analyzing Information Transfer from Heterogeneous Sources via Precise High-dimensional Asymptotics

### 2021-07-23 revision_journal/results.tex
intro wip
start sec 1.2
finish sec 1.2, begin sec 1.3

\section{Main results}\label{sec_main}

@@

\begin{figure}[!t]
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=0.8\textwidth]{figures/verify_covariate_shift.eps}
		\caption{Illustrating variance estimates for finite $p$}
		\label{fig_sec3_verify_cov}
	\end{subfigure}
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=0.8\textwidth]{figures/covariate_shift.eps}
		\caption{Varying the covariate shift parameter $\lambda$}
		\label{fig_sec3_covariate}
	\end{subfigure}
	\caption{We verify that Theorem \ref{thm_main_RMT} provides incredibly accurate estimates of the empirical variance under various finite sample sizes and dimensions.
	Furthermore, we illustrate an intriguing dichotomy between sample sizes and covariate (cf. Claim \ref{claim_dichotomy}).
	When $n_1 \ge n_2$, the lowest risk for predicting task two's labels is achieved by transferring from a dataset that has the same covariance matrix as task two. 
	When $n_1 < n_2$, the lowest risk is instead achieved by transferring a covariate-shifted dataset.
	Both simulation uses $\sigma = 1/2$.
	Figure \ref{fig_sec3_verify_cov} fixes $\lambda = 4$ and varies $n_1, p$.
	Figure \ref{fig_sec3_covariate} fixes $p = 100, n_2 = 300$ while varying $n_1, \lambda$.}
	\label{fig_sec31}
\end{figure}

@@

    We first describe a setting for modeling covariate shift.
    Let $\cS$ be the set of positive definite matrices $M^{\top} M$ such that for $i = 1,2,\dots,\floor{\frac p 2}$, $\lambda_{p - i}  = \frac 1 {\lambda_i}$, where $\lambda_1 \ge \lambda_2 \ge \cdots \ge \lambda_p$ are the eigenvalues of $M$.
    Under this setting, we ask: suppose that the sample sizes $n_1$ and $n_2$ are both fixed, which $M$ provides the best ``transfer'' in the sense of HPS estimator's excess risk?
    We observe a dichotomy that depends on whether or not $n_1$ is greater than $n_2$.
    \begin{claim}[Sample sizes vs. covariate shift]\label{claim_dichotomy}
        Let $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
        Suppose that $n_1$ and $n_2$ are both fixed.
        Within the set of all possible covariate shift matrix $M\in\cS$, the following dichotomy holds:
        \begin{enumerate}
	        \item[i)] If $n_1 \ge n_2$, then $g(M)$ is minimized in $\cS$ when $\lambda_i = 1$, for any $i = 1,\dots,p$.
	        \item[ii)] If $n_1 < n_2$, then $g(M)$ is maximized in $\cS$ when $\lambda_i = 1$, for any $i = 1,\dots,p$.
        \end{enumerate}
    \end{claim}

@@

Our second example illustrates that the effect of covariate shift worsens as the sample size of task one increases.
We consider a set of covariate shift matrices $M^{\top} M$ whose determinant are all equal to one, and the eigenvalues are bounded between $\tau^2$ and $1 / \tau^2$.
Let $\cS$ denote the set of $M\in\real^{p\times p}$ such that $M^{\top} M$ satisfies both conditions.
\begin{claim}[Identical covariance provides approximately optimal transfer under imbalanced dataset sizes]\label{prop_covariate}
    Recall that $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
	We have the following result regarding $g(\id_{p \times p})$:
	\begin{align} g(\id_{p\times p}) \le \bigbrace{1+ {\frac{n_2}{\tau^2 n_1}  }} g(M), \text{ for any } M \in \cS. \label{eq_claim_id}
	\end{align}
\end{claim}
\begin{proof}
    We can write the trace of $(\alpha_1 M^{\top} M + \alpha_2)^{-1}$ using the eigenvalues of $M^{\top} M$ as follows:
    \begin{align*}
        \bigtr{(\alpha_1 M^{\top} M + \alpha_2)^{-1}} &= \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2} \\
        &\ge \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2 \cdot {\lambda_i^2} / {\tau^2}} \tag{since $\lambda_i \ge \tau$, for any $i$} \\
        &= \frac 1 {\alpha_1 + \alpha_2 / \tau^2} \sum_{i=1}^ p \frac 1 {\lambda_i^2} \\
        &\ge \frac 1 {\alpha_1 + \alpha_2 /\tau^2} p\cdot \bigbrace{\prod_{i=1}^p \frac 1 {\lambda_i^2}}^{1/ p} \tag{by the AM-GM inequality} \\
        &= \frac{p}{\alpha_1 + \alpha_2 / \tau^2} \tag{since $\prod_{i=1}^p \lambda_i^2 = \det(M^{\top} M) = 1$}
    \end{align*}
    Next, recall from equation \eqref{g_id} that
    $g(\id_{p\times p}) = \frac{\sigma^2 p}{n_1 + n_2 - p}$.
    Thus, equation \eqref{eq_claim_id} follows if we show that (after rearranging terms)
    \begin{align}
        \frac{n_1 + n_2}{n_1 + n_2 - p} \cdot \frac 1 {\alpha_1 + \frac{\alpha_2}{\tau^2}} \le 1 + \frac{n_2}{\tau^2 \cdot n_1}. \label{eq_claim_id_step}
    \end{align}
    From the second part of equation \eqref{eq_a12extra}, we have that
    \begin{align*}
        \frac{n_1} {n_1 + n_2} &= \alpha_1 + \frac 1 {n_1 + n_2} \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2\alpha_1}{\lambda_i^2\alpha_1 + \alpha_2}} \\
        &< \alpha_1 + \frac p {n_1 + n_2}.
    \end{align*}
    Thus, $\alpha_1 > \frac{n_1 - p}{n_1 + n_2}$, which implies that $\alpha_2 < \frac{n_2}{n_1 + n_2}$.
    Hence, $\alpha_1 + \frac{\alpha_2}{\tau^2} = 1 - \frac{p}{n_1 + n_2} - \alpha_2 + \frac{\alpha_2}{\tau^2} \le 1 - \frac{p}{n_1 + n_2} + \frac{n_2}{n_1 + n_2} \frac 1 {\tau^2}$, which implies
    equation \eqref{eq_claim_id_step} by straightforward calculations.
    Thus, the proof is complete.
\end{proof}
As a corollary of Claim \ref{prop_covariate}, when $n_1$ is much larger than $n_2 / \tau^2$, the excess risk of HPS is approximately optimal in the set of all possible $M\in\cS$ when there is no covariate shift between dataset one and two.

@@

\begin{claim}[Sample sizes vs. model shift]\label{claim_model_shift}
    Suppose Assumption \ref{assm_big1} and the random-effect model under equation \eqref{eq_re} and \eqref{para_rel} holds.
    Suppose further that $n_2 \ge 3p$.
    Then, there exists a large constant $C > 0$ such that the following holds w.h.p.,
    \begin{enumerate}
        \item[i)] If $\mu^2 \le \frac{\sigma^2 p}{2(n_2 - p)}$, then $L(\hat{\beta}_2^{\MTL}) \le L(\hat{\beta}_2^{\STL}) + \OO(p^{-C})$.
        \item[ii)] If $\frac{\sigma^2 p}{2(n_2 - p)} < \mu^2 < \frac{\sigma^2 n_2}{2(n_2 - p)}$, then there exists a deterministic constant $\rho$ such that if $n_1 \le \rho\cdot p$, then $L(\hat{\beta}_2^{\MTL}) \le L(\hat{\beta}_2^{\STL}) + \OO(p^{-C})$, else $L(\hat{\beta}_2^{\STL}) \le L(\hat{\beta}_2^{\MTL}) + \OO(p^{-C})$.
        \item[iii)] If $\frac{\sigma^2 n_2} {2(n_2 - p)} \le \mu^2$, then $L(\hat{\beta}_2^{\STL}) \le L(\hat{\beta}_2^{\MTL}) + \OO(p^{-C})$.
    \end{enumerate}
\end{claim}

>>>>

\section{Precise asymptotics under distribution shift}\label{sec_main}

@@

\begin{figure}[!t]
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=0.8\textwidth]{figures/verify_covariate_shift.eps}
		\caption{Illustrating HPS estimates in finite dimensions}
		\label{fig_sec3_verify_cov}
	\end{subfigure}
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=0.8\textwidth]{figures/covariate_shift.eps}
		\caption{Varying the covariate shift parameter $\lambda$}
		\label{fig_sec3_covariate}
	\end{subfigure}
	\caption{We verify that Theorem \ref{thm_main_RMT} provides incredibly accurate estimates of the empirical variance under various finite sample sizes and dimensions.
	Furthermore, we illustrate an intriguing dichotomy between sample sizes and covariate (cf. Claim \ref{claim_dichotomy}).
	When $n_1 \ge n_2$, the lowest risk for predicting task two's labels is achieved by transferring from a dataset that has the same covariance matrix as task two. 
	When $n_1 < n_2$, the lowest risk is instead achieved by transferring a covariate-shifted dataset.
	Both simulation uses $\sigma = 1/2$.
	Figure \ref{fig_sec3_verify_cov} fixes $\lambda = 4$ and varies $n_1, p$.
	Figure \ref{fig_sec3_covariate} fixes $p = 100, n_2 = 300$ while varying $n_1, \lambda$.}
	\label{fig_sec31}
\end{figure}

@@

    We first describe a setting for modeling covariate shift.
    Let $\cS$ be the set of positive definite matrices $M^{\top} M$ such that for $i = 1,2,\dots,\floor{\frac p 2}$, $\lambda_{p - i}  = \frac 1 {\lambda_i}$, where $\lambda_1 \ge \lambda_2 \ge \cdots \ge \lambda_p$ are the eigenvalues of $M$.
    Under this setting, we ask: suppose that the sample sizes $n_1$ and $n_2$ are both fixed, which $M$ provides the best ``transfer'' in the sense of HPS estimator's excess risk?
    We observe a dichotomy that depends on whether or not $n_1$ is greater than $n_2$.
    \begin{proposition}[Sample sizes vs. covariate shift]\label{claim_dichotomy}
        Let $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
        Suppose that $n_1$ and $n_2$ are both fixed.
        Within the set of all possible covariate shift matrix $M\in\cS$, the following dichotomy holds:
        \begin{enumerate}
	        \item[i)] If $n_1 \ge n_2$, then $g(M)$ is minimized in $\cS$ when $\lambda_i = 1$, for any $i = 1,\dots,p$.
	        \item[ii)] If $n_1 < n_2$, then $g(M)$ is maximized in $\cS$ when $\lambda_i = 1$, for any $i = 1,\dots,p$.
        \end{enumerate}
    \end{proposition}

@@

Our second example illustrates that the effect of covariate shift worsens as the sample size of task one increases.
We consider a set of covariate shift matrices $M^{\top} M$ whose determinant are all equal to one, and the eigenvalues are bounded between $\tau^2$ and $1 / \tau^2$.
Let $\cS$ denote the set of $M\in\real^{p\times p}$ such that $M^{\top} M$ satisfies both conditions.
\begin{proposition}[Identical covariance provides approximately optimal transfer under imbalanced dataset sizes]\label{prop_covariate}
    Recall that $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
	We have the following result regarding $g(\id_{p \times p})$:
	\begin{align} g(\id_{p\times p}) \le \bigbrace{1+ {\frac{n_2}{\tau^2 n_1}  }} g(M), \text{ for any } M \in \cS. \label{eq_claim_id}
	\end{align}
\end{proposition}
\begin{proof}
    We can write the trace of $(\alpha_1 M^{\top} M + \alpha_2)^{-1}$ using the eigenvalues of $M^{\top} M$ as follows:
    \begin{align*}
        \bigtr{(\alpha_1 M^{\top} M + \alpha_2)^{-1}} &= \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2} \\
        &\ge \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2 \cdot {\lambda_i^2} / {\tau^2}} \tag{since $\lambda_i \ge \tau$, for any $i$} \\
        &= \frac 1 {\alpha_1 + \alpha_2 / \tau^2} \sum_{i=1}^ p \frac 1 {\lambda_i^2} \\
        &\ge \frac 1 {\alpha_1 + \alpha_2 /\tau^2} p\cdot \bigbrace{\prod_{i=1}^p \frac 1 {\lambda_i^2}}^{1/ p} \tag{by the AM-GM inequality} \\
        &= \frac{p}{\alpha_1 + \alpha_2 / \tau^2} \tag{since $\prod_{i=1}^p \lambda_i^2 = \det(M^{\top} M) = 1$}
    \end{align*}
    Next, recall from equation \eqref{g_id} that
    $g(\id_{p\times p}) = \frac{\sigma^2 p}{n_1 + n_2 - p}$.
    Thus, equation \eqref{eq_claim_id} follows if we show that (after rearranging terms)
    \begin{align}
        \frac{n_1 + n_2}{n_1 + n_2 - p} \cdot \frac 1 {\alpha_1 + \frac{\alpha_2}{\tau^2}} \le 1 + \frac{n_2}{\tau^2 \cdot n_1}. \label{eq_claim_id_step}
    \end{align}
    From the second part of equation \eqref{eq_a12extra}, we have that
    \begin{align*}
        \frac{n_1} {n_1 + n_2} &= \alpha_1 + \frac 1 {n_1 + n_2} \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2\alpha_1}{\lambda_i^2\alpha_1 + \alpha_2}} \\
        &< \alpha_1 + \frac p {n_1 + n_2}.
    \end{align*}
    Thus, $\alpha_1 > \frac{n_1 - p}{n_1 + n_2}$, which implies that $\alpha_2 < \frac{n_2}{n_1 + n_2}$.
    Hence, $\alpha_1 + \frac{\alpha_2}{\tau^2} = 1 - \frac{p}{n_1 + n_2} - \alpha_2 + \frac{\alpha_2}{\tau^2} \le 1 - \frac{p}{n_1 + n_2} + \frac{n_2}{n_1 + n_2} \frac 1 {\tau^2}$, which implies
    equation \eqref{eq_claim_id_step} by straightforward calculations.
    Thus, the proof is complete.
\end{proof}
As a corollary of Claim \ref{prop_covariate}, when $n_1$ is much larger than $n_2 / \tau^2$, the excess risk of HPS is approximately optimal in the set of all possible $M\in\cS$ when there is no covariate shift between dataset one and two.

@@

\begin{proposition}[Sample sizes vs. model shift]\label{claim_model_shift}
    Suppose Assumption \ref{assm_big1} and the random-effect model under equation \eqref{eq_re} and \eqref{para_rel} holds.
    Suppose further that $n_2 \ge 3p$.
    Then, there exists a large constant $C > 0$ such that the following holds w.h.p.,
    \begin{enumerate}
        \item[i)] If $\mu^2 \le \frac{\sigma^2 p}{2(n_2 - p)}$, then $L(\hat{\beta}_2^{\MTL}) \le L(\hat{\beta}_2^{\STL}) + \OO(p^{-C})$.
        \item[ii)] If $\frac{\sigma^2 p}{2(n_2 - p)} < \mu^2 < \frac{\sigma^2 n_2}{2(n_2 - p)}$, then there exists a deterministic constant $\rho$ such that if $n_1 \le \rho\cdot p$, then $L(\hat{\beta}_2^{\MTL}) \le L(\hat{\beta}_2^{\STL}) + \OO(p^{-C})$, else $L(\hat{\beta}_2^{\STL}) \le L(\hat{\beta}_2^{\MTL}) + \OO(p^{-C})$.
        \item[iii)] If $\frac{\sigma^2 n_2} {2(n_2 - p)} \le \mu^2$, then $L(\hat{\beta}_2^{\STL}) \le L(\hat{\beta}_2^{\MTL}) + \OO(p^{-C})$.
    \end{enumerate}
\end{proposition}

### 2021-07-23 revision_journal/setup.tex
start sec 1.2
finish sec 1.2, begin sec 1.3

\section{Preliminaries}\label{sec_HPS}
We begin by formally defining the data model involving large dimensional random matrices together with the underling assumptions.
We define the covariate shift and model shift settings.
Then, we describe the HPS estimators for these settings using two-layer linear neural networks.
Finally, we present a bias-variance decomposition of the HPS estimators.
We explain why this decomposition is crucial for analyzing information transfer and the need for new techniques in the analysis.
Our setup in this section concerns the case of a single data source.
Later on in Section \ref{sec_same}, we will extend the setup to the case of multiple data sources.
%In Section \ref{sec_same}, we will consider a simple extension to a multi-task learning setting with more than two tasks and  the same covariates for all tasks. We will give the asymptotic limit of the predication loss for the HPS estimator, which provides useful insights into the more rich multi-task learning theory. %Due to length constraint, we will explore the multi-task learning setting in greater details in other works.

@@

\subsection{Risks}

@@

 %The above result has a long history in random matrix theory.
If the entries of $Z^{(2)}$ are i.i.d. Gaussian, then this result follows from the classical result for %the mean of
the inverse Wishart distribution \citet{anderson1958introduction}. For a general non-Gaussian $Z^{(2)}$, this result can be obtained using the well-known Stieltjes transform method (cf. Lemma 3.11 of \citet{bai2009spectral}). Here we have presented the results from \citet{isotropic} and \citet{DY}, who give an almost sharp convergence rate.

On the other hand, characterizing the risk of the HPS estimator requires dealing with several challenges.
First, characterizing bias (cf. equation \eqref{Lbias}) introduces the issue of \textit{model shift}, when $\beta^{(1)}$ differs from $\beta^{(2)}$.
Second, characterizing variance (cf. equation \eqref{Lvar}) introduces  the issue of \textit{covariate shift}, when $\Sigma^{(1)}$ differs from $\Sigma^{(2)}$.
Our main results in the next section provide precise bias-variance asymptotics under (either or both) covariate and model shifts.

>>>>

\section{Preliminaries}\label{sec_HPS}
We begin by formally defining the data model involving large dimensional random matrices together with the underling assumptions.
We define the covariate shift and model shift settings.
Then, we describe the HPS estimators for these settings using two-layer linear neural networks.
Finally, we present a bias-variance decomposition of the HPS estimators.
We explain why this decomposition is crucial for analyzing information transfer and the need for new techniques in the analysis.
Our setup in this section concerns the case of a single data source.
Later on in Section \ref{sec_same}, we will extend the setup to the case of multiple data sources.

@@

\subsection{Risks}\label{sec_risk}

@@

 %The above result has a long history in random matrix theory.
If the entries of $Z^{(2)}$ are i.i.d. Gaussian, then this result follows from the classical result for %the mean of
the inverse Wishart distribution \cite{anderson1958introduction}. For a general non-Gaussian $Z^{(2)}$, this result can be obtained using the well-known Stieltjes transform method (cf. Lemma 3.11 of \citet{bai2009spectral}). Here we have presented the results from \citet{isotropic} and \citet{DY}, who give an almost sharp convergence rate.
